p-adic stability in linear algebra

### 2015-01-17 examples/stability.tex
Updating charpoly section, fixing typos in A.1

\newcommand{\noopsort}[1]{}
\DeclareMathOperator{\NP}{NP}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\val}{val}
\DeclareMathOperator{\pr}{pr}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\com}{Com}
\DeclareMathOperator{\Grass}{Grass}
\DeclareMathOperator{\Lat}{Lat}
\DeclareMathOperator{\round}{round}

@@

\noindent
{\bf Notations.}
Throughout the paper, the letter $K$ will refer to a \emph{complete 
discrete valuation field of characteristic $0$}, \todo{why characteristic 0?} that is either a finite 
extension of the field of $p$-adic numbers $\Qp$ or a field of Laurent 
series over a field of characteristic $0$. We denote by $\val : K \to \Z 
\cup \{+\infty\}$ of $K$ --- it is the $p$-adic valuation when $K = \Qp$ 
and the usual valuation of a Laurent series when $K = k((t))$ --- and by 
$\Vert \cdot \Vert$ the corresponding norm.

@@

We assume that we are given in addition two nondecreasing convex 
functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We suppose further that 
there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
$({-}\infty, \nu]$\footnote{We note that this assumption is fullfilled if 
we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
does not vanish.}. We introduce the functions $\tau_\nu$ and $\Lambda_f$ 
defined by:
$$\begin{array}{rll}
&\tau_\nu(x) = x & \text{if } x \leq \nu \\
& \hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise} \medskip \\
\text{and} &
\multicolumn{2}{l}{\Lambda_f(x) = 
  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)}
\end{array}$$
where $\alpha$ is a real number satisfying $\Vert n! \Vert \geq 
e^{-\alpha n}$ for all $n$. A suitable value for $\alpha$ is $\alpha = - 
\frac p {p-1} \cdot \log \Vert p \Vert > 0$ where $p$ denotes the 
characteristic of the residue field and where, by convention, the above 
expression equals $0$ if $p = 0$. \todo{which expression; the one defining $\alpha$?}
The next Proposition is proved in Appendix \ref{app:proof}.

\begin{prop}
\label{prop:boundLambdaf}
We have $\Lambda(f) \leq \Lambda_f$.
\end{prop}

@@

Figure \ref{fig:area} illustrates Proposition \ref{prop:boundLambdaf}. 
The blue \todo{do we have color?} plain line represents the graph of the function $\Lambda_f$. A 
quick computation shows that, on a neighborhood of ${-}\infty$, this 
function is given by $\Lambda_f(x) = x + \alpha + \mu$
where $\mu$ is the value that $\Lambda_g$ takes on the interval 
$({-}\infty, \nu]$. Proposition \ref{prop:boundLambdaf} says that the
graph of $\Lambda(f)$ lies below the plain blue line. We remark
moreover that the Taylor expansion of $f(z)$ starts with the term
$g(0) z$. Hence, on a neighborhood on ${-}\infty$, we have 
$\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
get:
$$\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert, 
  \quad \forall x \in \R.$$
In other words, the graph of $\Lambda(f)$ lies above the brown line.
Furthermore, we know that the slopes of $\Lambda(f)$ are all integral
because $f$ is locally analytic. Hence, $\Lambda(f)$ cannot lie above
the dashed blue line defined as the line of slope $2$ passing through
the first break point of the blue plain line --- which has coordinate 
$(y_0 - \alpha - \mu, y_0)$ with $y_0 = \min(\Lambda_h^{-1}(\nu) + \mu, 
\nu)$. As a conclusion, we have proved that the graph of $\Lambda(f)$ 
must coincide with the brown line until it meets the dashed blue line 
and then has to stay in the green area.

@@

\todo{Write a short introduction.}

The notation $M_{n,m}(K)$ refers to the $K$-vector space of matrices 
with $n$ rows and $m$ columns with coefficients in $K$.

@@

To fix ideas, let us assume from now that the entries of $A$ and $B$ all 
lie in $\O_K$ and are known at the same precision $O(\pi^N)$. In order 
to apply Propositions \ref{prop:precision} and \ref{prop:precision2}, we then need to compute the image 
of the standard lattice $\mathcal L_0 = M_{r,s}(\O_K) \times 
M_{s,t}(\O_K)$ under $\mathcal P'_{r,s,t}(A,B)$. It is of course 
contained in $M_{r,t}(\O_K)$; this reflects the obvious fact that each 
entry of the product $AB$ is also known with precision at least $O(\pi^N)$. 
Nonetheless, it may happen that the above inclusion is strict, meaning 
that we are \emph{gaining} precision in those cases. In order to compute 
exactly $\mathcal P'_{a,b,c}(A,B)(\mathcal L_0)$, we introduce the Smith 
factorization of $A$ and $B$:
$$A = U_A \cdot \Delta_A \cdot V_A
\text{ and } B = U_B \cdot \Delta_B \cdot V_B$$
and denote by $a_1, \ldots, a_{\min(r,s)}$ (resp. $b_1, \ldots, 
b_{\min(s,t)}$) the valuation of the diagonal entries of $\Delta_A$
(resp. $\Delta_B$). We also agree to set $a_n = b_m = +\infty$ if
$n > \min(r,s)$ and $m > \min(s,t)$. We define $M_{r,t}((a_i),(b_j))$ 
as the sublattice of $M_{r,t}(\O_K)$ consisting of matrices $M = (M_{i,j})$ 
such that $\val(M_{i,j}) \geq \min(a_i,b_j)$ for all $(i,j)$.

\begin{prop}
\label{prop:mulmatrix}
With the above notations, we have:
$$\begin{array}{rl}
& \mathcal P'_{r,s,t}(A,B)(\mathcal L_0)
= U_A \cdot M_{r,t}((a_i),(b_j)) \cdot V_B \smallskip \\
\text{and} &
\text{\emph{length}}\Big(\frac{M_{r,t}(\O_K)}{\mathcal P'_{r,s,t}(A,B)(\mathcal L_0) }
\Big) =
\displaystyle 
\sum_{m=1}^\infty a^{(m)} b^{(m)}
\end{array}$$
where $a^{(m)}$ (resp. $b^{(m)}$) is the number of $a_i$'s (resp.
of $b_j$'s) which are $\geq m$.
\end{prop}

@@

As we did for matrix multiplication, we first determine the image of
the standard lattice $\mathcal L_0 = M_{n,n}(\O_K)$ under $\detp(M)$.
In order to do so, we write $M = U_M \cdot \Delta_M \cdot V_M$ the 
Smith factorization of $M$ and denote by $m_1 \leq \ldots \leq m_n$ 
the valuation of the diagonal entries of $\Delta_M$.

\begin{prop}
\label{prop:detmatrix}
Setting $v = m_1 + \cdots + m_{n-1}$, we have
$\detp(M)(\mathcal L_0) = \pi^v \O_K$.
\end{prop}

\begin{proof}
From the description of $\detp(M)$, we see that it is enough to prove 
that the smallest valuation of an entry of $\com(M)$ is $v$
or, equivalently, that the ideal of $\O_K$ generated by all 
minors of $M$ of size $(n-1)$ is $\pi^v
\O_K$. But this ideal remains unchanged when we multiply $M$ 
(on the left or on the right) by any unimodular matrix. Thus we may 
assume that $M$ is $\Delta_M$ and the result becomes clear.
\end{proof}

In terms of precision, Proposition \ref{prop:detmatrix} implies that if 
$M$ is given at flat precision $O(\pi^N)$ with $N > v$, then $\det(M)$ 
is known at precision $O(\pi^{N+v})$, so we are gaining $v$ significant 
digits in absolute precision\footnote{We note that the valuation of 
$\det(M)$ is $v + m_n$; thus if we are considering \emph{relative} 
precisions then we should say that we are loosing $m_n$ significant 
digits.}.
Furthermore, there exists a easy method for computing the determinant
of $M$ with this optimal precision. It consists in first computing an
approximate Smith decomposition of $M$, that is a writing
$M = \tilde U_M \cdot \tilde \Delta_M \cdot \tilde V_M$
where $\tilde U_M$ and $\tilde V_M$ are exact unimodular matrices and
$\tilde \Delta_M$ is known at precision $O(\pi^N)$, is diagonal modulo
$\pi^N$ and has diagonal entries $a_1, \ldots, a_n$ of valuations 
$m_1, \ldots, m_n$ respectively.
We then have $\det(M) = \det(\tilde \Delta_M)$ and factoring $a_i$ on
the $i$-th row of $\tilde \Delta_M$, we obtain
$\det(\tilde \Delta_M) = a_1 \cdots a_m \cdot U$
where $U$ is a matrix which is congruent to the identity modulo 
$\pi^{N-m_n}$. Therefore we get:
$$\det(M) \equiv a_1 \cdots a_m \pmod{\pi^{N+v}}$$
from what we see that we can compute $\det(M)$ at the required precision
by simply evaluating the product $a_1 \cdots a_m$.

\subsection{Characteristic polynomials}

@@

For any precision lattice $H$, the precision of the characteristic polynomial will thus be the lattice in $K_{<n}[X]$ generated by the polynomials $\tr(\com(X - M) \cdot dM)$, as $dM$ ranges over a set of generators for $H$.  In the standard case that the precision lattice for $M$ is just $H = \pi^N M_{n,n}(\O_K)$, we can further analyze $\charp(H)$.

Recall that the \emph{Newton polygon} of a polynomial $\sum_{i} a_i X^i$ is the lower convex hull of the points $(i, \val(a_i))$ and the \emph{Newton polygon} of a matrix is the Newton polygon of its characteristic polynomial.  If $m_1 \le \cdots \le m_n$ are the valuations of the diagonal entries of $\Delta_M$ in a Smith factorization $U_M \cdot \Delta_M \cdot V_M$ of $M$ as in the previous section, the \emph{Hodge polygon} of $M$ is the lower convex hull of the points $(i, \sum_{j = i+1}^n m_i)$.

\begin{deftn}
The \emph{precision polygon} of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X - M)$.  
\end{deftn}

Any such polygon $P$ naturally yields a lattice in $K_{<n}[X]$, consisting of polynomials whose Newton polygons lie above $P$.  The lattice associated to $P$ is generated by monomials $a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.

@@

\begin{prop}
Suppose a matrix $M$ is known with precision $H = \pi^N M_{n,n}(\O_K)$.  Then the precision of the characteristic polynomial of $M$ is contained within $S_N$ applied to the lattice of the precision polygon of $M$.
\end{prop}
\begin{proof}
The Newton polygon of each entry of $\com(X - M)$ lies above the precision polygon by definition.  Since these generate $\charp(M)(H)$, the result follows.
\end{proof}

\begin{prop} \label{prop:prec_hodge}
The precision polygon of $M$ lies above $T_1$ applied to the Hodge polygon of $M$.
\end{prop}

\begin{proof}
\todo{Add this proof from Xavier's e-mail}
\end{proof}

For many matrices, the precision polygon equals the Hodge polygon.  For integer matrices with entries uniformly chosen between $0$ and $1024$, the $2$-adic precision polygon is equal to the Hodge polygon in $99.5\%$ of cases in dimension $4$, down to $99.1\%$ in dimension $8$.  \todo{p=3?}  Moreover, experiments support the following conjectures.

\begin{conj}
\begin{enumerate}
\item The precision polygon is equal to the Hodge polygon if the determinant of $M$ has valuation $0$ or $1$, or if $M$ is $3 \times 3$.
\item The height at $0$ of the precision polygon is equal to the height at $1$ of the Hodge polygon.
\item The precision polygon lies below $T_1$ applied to the Newton polygon.
\end{enumerate}
\end{conj}
Empirically, the precision polynomial seems most likely to differ from the Hodge polygon at $1$, corresponding to the precision of the linear term of the characteristic polynomial.

Of course, the precision polygon does necessarily agree with the full precision of the characteristic polynomial of $M$.  In $11\%$ of cases in dimension $3$, up to $15\%$ of cases in dimension $8$, the lattice associated to the precision polygon contains the actual precision lattice with index greater than $1$.  This percentage increases as the valuation of the determinant increases, reaching $34\%$ in dimension $9$ for matrices constrained to have determinant with $2$-adic valuation $12$.

Moreover, one can construct examples with arbitrarily large extra lattice precision.  If $M$ has elementary divisors with valuations $m_1 \le \cdots \le m_n$, Proposition \ref{prop:prec_hodge} implies that the precision lattice of the characteristic polynomial of $M$ is contained within $\O_{K,n}[X]$ with index at least $\sum_{i=1}^{n-1} m_i$.  The precision lattice of $1 + M$ is obtained from the precision lattice of $M$ via the transformation $X \mapsto X + 1$, but the precision polygon is now flat with height $0$.

@@

\noindent
{\bf Numerical experiments.}
Let $B_n=(E_{i,j})_{1 \leq i,j\leq n}$ be the canonical basis of 
$M_n(K)$. It can be naturally seen as a basis of $L_n \times U_n$ as
well. For a given $M \in O_n$, let us abuse notation and write $D'(M)$
for the matrix of this linear mapping in the above basis. We remark that
$D'(M)$ is lower-triangular in this basis. Projecting $D'(M)$ onto each
coordinate, we find the best coordinate-wise loss of precision we can 
hope for the computation of $D$ is given by $\sum_u \left( \max_v \left( 
\val(D'(M)_{u,v}) \right) \right)$. This number should be compared to
$\val(\det(D'(M)))$, which is the lattice-method loss of precision.
Figure \ref{fig:LU} sums up the mean and standard deviation of those 
losses for a sample of 2000 random matrices in $M_d(\mathbb{Z}_2)$.
%
\begin{figure}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|c|c|c|c|}
\hline 
& \multicolumn{4}{|c|}{Loss of precision in LU decomposition} \\
\cline{2-5}
\raisebox{0.2em}{matrix}
& \multicolumn{2}{|c|}{coord-wise method} & \multicolumn{2}{|c|}{lattice method}  \\  \cline{2-5}
\smash{\raisebox{0.6em}{size}} 
& \hspace{0.5em}mean\hspace{0.5em} & deviation 
& \hspace{0.5em}mean\hspace{0.5em} & deviation \\ \hline
$2$ & $\phantom{2}{-}3.0$& $5$ & $-1.5$ & $1.4$ \\% \hline
$3$& $\phantom{2}{-}9.4$& $11$ & $-2.3$ & $2.3$\\% \hline
$4$ & $-20\phantom{.4}$& $20$ & $-3.8$ & $3.1$\\ \hline
\end{tabular} 
\smallskip

@@

When the couple $(V, V^\c)$ varies, the open subsets $\mathcal 
U_{V,V^\c}$ cover the whole Grassmaniann and then define an atlas 
of it. When implementing vector spaces on computer, we usually restrict 
ourselves to the subatlas consisting of all charts of the form $(V_I, 
V_{I^\c})$ where $I$ runs over the family of subsets of $\{1, 
\ldots, n\}$ of cardinality $d$ and $V_I$ is the subspaces spanned by 
the $e_i$'s with $i \in I$. A subspace $W \in E$ then belongs to at 
least one $\mathcal U_{V_I, V_{I^{\text{c}}}}$ and, given a family of
generators of $W$, we can determine such an $I$ together with the
corresponding embedding $f : V_I \hookrightarrow E$ by echelonizing 
the matrix of generators of $W$.

@@

\noindent
{\bf A variant.}
Alternatively, one can describe $\Grass(E,d)$ as the set of linear 
surjective morphisms $f : E \to E/V$ modulo the action (by 
postcomposition) of $\GL(E/V)$ as well. This identification presents the 
tangent space at a given point $V$ as the quotient $\Hom(V, E/V) / 
\End(E/V)$, which turns out to be again isomorphic to $\Hom(V, E/V)$. 
Given a decomposition $E = V \oplus V^\c$ as above, we let $\mathcal 
U^\star_{V, V^\c}$ denote the set of surjective linear maps $f : E 
\to V^\c$ whose restriction to $V^\c$ is the identity. It is an 
affine space over $\Hom(V, V^\c)$ which can be identified to an open
subset of $\Grass(E,d)$ \emph{via} the map $f \mapsto \ker f$.

It is actually easily seen that $\mathcal U_{V, V^\c}$ and $\mathcal 
U^\star_{V, V^\c}$ define the same open subset in $\Grass(E,d)$. 
Indeed, given $f \in \mathcal U_{V, V^\c}$, one can write $f = \id_V 
+ h$ with $h \in \Hom(V, V^\c)$ and define the morphism $g = \id_E -
h \circ \pi \in \mathcal U^\star_{V, V^\c}$. The association $f \mapsto
g$ then defines a bijection $\mathcal U_{V, V^\c} \to \mathcal
U^\star_{V, V^\c}$ which commutes with the embeddings into the
Grassmannian.

>>>>

\newcommand{\noopsort}[1]{}
\DeclareMathOperator{\NP}{NP}
\DeclareMathOperator{\HP}{HP}
\DeclareMathOperator{\PP}{PP}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\val}{val}
\DeclareMathOperator{\pr}{pr}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\com}{Com}
\DeclareMathOperator{\Grass}{Grass}
\DeclareMathOperator{\Lat}{Lat}
\DeclareMathOperator{\round}{round}

@@

\noindent
{\bf Notations.}
Throughout the paper, the letter $K$ will refer to a complete 
discrete valuation field of characteristic $0$, that is either a finite 
extension of the field of $p$-adic numbers $\Qp$ or a field of Laurent 
series over a field of characteristic $0$ \todo{what about mixed characteristic with infinite residue field?}.
We denote by $\val : K \to \Z 
\cup \{+\infty\}$ of $K$ --- it is the $p$-adic valuation when $K = \Qp$ 
and the usual valuation of a Laurent series when $K = k((t))$ --- and by 
$\Vert \cdot \Vert$ the corresponding norm.

@@

We assume that we are given in addition two nondecreasing convex 
functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We suppose further that 
there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
$({-}\infty, \nu]$\footnote{We note that this assumption is fullfilled if 
we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
does not vanish.}. We introduce the functions $\tau_\nu$ and $\Lambda_f$ 
defined by:
$$\begin{array}{rll}
&\tau_\nu(x) = x & \text{if } x \leq \nu \\
& \hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise} \medskip \\
\text{and} &
\multicolumn{2}{l}{\Lambda_f(x) = 
  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)}
\end{array}$$
where $\alpha$ is a real number satisfying $\Vert n! \Vert \geq 
e^{-\alpha n}$ for all $n$. If $p$ is the characteristic of the residue field,
a suitable value for $\alpha$ is $\alpha = - 
\frac p {p-1} \cdot \log \Vert p \Vert$ if $p > 0$ and $\alpha = 0$ if $p = 0$.
The next Proposition is proved in Appendix \ref{app:proof}.

\begin{prop}
\label{prop:boundLambdaf} \label{PROP:BOUNDLAMBDAF}
We have $\Lambda(f) \leq \Lambda_f$.
\end{prop}

@@

Figure \ref{fig:area} illustrates Proposition \ref{prop:boundLambdaf}. 
The blue plain line represents the graph of the function $\Lambda_f$. A 
quick computation shows that, on a neighborhood of ${-}\infty$, this 
function is given by $\Lambda_f(x) = x + \alpha + \mu$
where $\mu$ is the value that $\Lambda_g$ takes on the interval 
$({-}\infty, \nu]$. Proposition \ref{prop:boundLambdaf} says that the
graph of $\Lambda(f)$ lies below the plain blue line. We remark
moreover that the Taylor expansion of $f(z)$ starts with the term
$g(0) z$. Hence, on a neighborhood on ${-}\infty$, we have 
$\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
get:
$$\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert, 
  \quad \forall x \in \R.$$
In other words, the graph of $\Lambda(f)$ lies above the brown line.
Furthermore, we know that the slopes of $\Lambda(f)$ are all integral
because $f$ is locally analytic. Hence, $\Lambda(f)$ cannot lie above
the dashed blue line defined as the line of slope $2$ passing through
the first break point of the blue plain line --- which has coordinate 
$(y_0 - \alpha - \mu, y_0)$ with $y_0 = \min(\Lambda_h^{-1}(\nu) + \mu, 
\nu)$. As a conclusion, we have proved that the graph of $\Lambda(f)$ 
must coincide with the brown line until it meets the dashed blue line 
and then has to stay in the green area.

@@

The notation $M_{m,n}(K)$ refers to the $K$-vector space of matrices 
with $m$ rows and $n$ columns with coefficients in $K$.  We will repeatedly
use the Smith decomposition for $M \in M_{m,n}(K)$:
\[
M = U_M \cdot \Delta_M \cdot V_M,
\]
with $U_M$ and $V_M$ unimodular and $\Delta_M$ diagonal.  Write $\sigma_i(M)$
for the valuation of the $(i,i)$-th entry of $\Delta_M$, and by convention set
$\sigma_i(M) = +\infty$ if $i > \min(m,n)$.  Order the $\sigma_i(M)$ so that $\sigma_i(M) \le \sigma_{i+1}(M)$.

@@

To fix ideas, let us assume from now that the entries of $A$ and $B$ all 
lie in $\O_K$ and are known at the same precision $O(\pi^N)$. In order 
to apply Propositions \ref{prop:precision} and \ref{prop:precision2}, we then need to compute the image 
of the standard lattice $\mathcal L_0 = M_{r,s}(\O_K) \times 
M_{s,t}(\O_K)$ under $\mathcal P'_{r,s,t}(A,B)$. It is of course 
contained in $M_{r,t}(\O_K)$; this reflects the obvious fact that each 
entry of the product $AB$ is also known with precision at least $O(\pi^N)$. 
Nonetheless, it may happen that the above inclusion is strict, meaning 
that we are \emph{gaining} precision in those cases.

Set $a_i = \sigma_i(A)$ and $b_i = \sigma_i(B)$, and define $M_{r,t}((a_i),(b_j))$ 
as the sublattice of $M_{r,t}(\O_K)$ consisting of matrices $M = (M_{i,j})$ 
such that $\val(M_{i,j}) \geq \min(a_i,b_j)$ for all $(i,j)$.  Write $a^{(m)}$ (resp. $b^{(m)}$)
for the number of $a_i$'s (resp. $b_i$'s) which are at least $m$.

\begin{prop}
\label{prop:mulmatrix}
With the above notations, we have
\[
\begin{array}{rl}
& \mathcal P'_{r,s,t}(A,B)(\mathcal L_0)
= U_A \cdot M_{r,t}((a_i),(b_j)) \cdot V_B \smallskip \\
\text{and} &
\text{\emph{length}}\Big(\frac{M_{r,t}(\O_K)}{\mathcal P'_{r,s,t}(A,B)(\mathcal L_0) }
\Big) =
\displaystyle 
\sum_{m=1}^\infty a^{(m)} b^{(m)}.
\end{array}
\]
\end{prop}

@@

As we did for matrix multiplication, we first determine the image of
the standard lattice $\mathcal L_0 = M_{n,n}(\O_K)$ under $\detp(M)$.

\begin{prop}
\label{prop:detmatrix}
Setting $v = \sigma_1(M) + \cdots + \sigma_{n-1}(M)$, we have
$\detp(M)(\mathcal L_0) = \pi^v \O_K$.
\end{prop}

\begin{proof}
From the description of $\detp(M)$, we see that it is enough to prove 
that the smallest valuation of an entry of $\com(M)$ is $v$
or, equivalently, that the ideal of $\O_K$ generated by all 
minors of $M$ of size $(n-1)$ is $\pi^v
\O_K$. But this ideal remains unchanged when we multiply $M$ 
on the left or on the right by a unimodular matrix. Thus we may 
assume that $M = \Delta_M$, and the result becomes clear.
\end{proof}

In terms of precision, Proposition \ref{prop:detmatrix} implies that if 
$M$ is given at flat precision $O(\pi^N)$ with $N > v$, then $\det(M)$ 
is known at precision $O(\pi^{N+v})$.  Thus we are gaining $v$ significant 
digits in absolute precision or, equivalently, losing $\sigma_n(M)$ digits
of relative precision.
Furthermore, one may easily compute the determinant
of $M$ with this optimal precision by finding an approximate
Smith decomposition with $\Delta_M$ known at precision $O(\pi^N)$ and
multiplying its diagonal entries.

\subsection{Characteristic polynomials} \label{ssec:charpoly}

@@

Write $\mathcal{E}$ for the precision lattice in $K_{<n}[X]$ of the characteristic polynomial of $M$, where $M$ is known with precision lattice $\mathcal{L}$.  As long as $\mathcal{L}$ is sufficiently small, Proposition \ref{prop:precision} implies that $\mathcal{E}$ will be generated by the polynomials $\tr(\com(X - M) \cdot dM)$, as $dM$ ranges over a set of generators for $\mathcal{L}$.  When $\mathcal{L} = \mathcal{L}_N := \pi^N M_{n,n}(\O_K)$, we can further analyze $\mathcal{E}$.

Recall that the \emph{Newton polygon} $\NP(f)$ of a polynomial $f(X) = \sum_{i} a_i X^i$ is the lower convex hull of the points $(i, \val(a_i))$ and the \emph{Newton polygon} $\NP(M)$ of a matrix $M$ is $\NP(\charpoly(M))$.  The \emph{Hodge polygon} of $M$ is the lower convex hull of the points $(i, \sum_{j = 1}^{n-i} \sigma_i(M))$.  Such polygons arise naturally in tracking the precision of polynomials \cite{caruso-roe-vaccon:14a}*{\S 4.2}.  Any such polygon $P$ yields a lattice $\mathcal{L}_P$ in $K_{<n}[X]$ consisting polynomials whose Newton polygons lie above $P$.  This lattice is generated by monomials $a_iX^i$, where $\val(a_i)$ is the ceiling of the height of $P$ at $i$.  One uses these polygons to do a certain kind of coordinate-wise precision tracking for polynomials, bounding the actual precision lattice.

For any matrix $M$, the Newton polygon lies above the Hodge polygon \cite{kedlaya:padicDiffEq}*{Thm. 4.3.11}.  We now introduce another polygon that will be bounded by between these two, and will provide an estimate for $\mathcal{E}$.

\begin{deftn}
The \emph{precision polygon} $\PP(M)$ of $M$ is the lower convex hull of the Newton polygons of the entries of $\com(X - M)$.  
\end{deftn}

@@

\begin{prop}
Suppose a matrix $M$ is known with precision $\mathcal{L}_N$ with $N$ sufficiently large.  Then $\mathcal{E} \subseteq \mathcal{L}_{S_N(\PP(M))}$.
\end{prop}
\begin{proof}
The Newton polygon of each entry of $\com(X - M)$ lies above the precision polygon by definition.  Since these generate $\charp(M)(\mathcal{L}_0)$, the result follows from Proposition \ref{prop:precision}.
\end{proof}

\begin{prop} \label{prop:prec_hodge}
The precision polygon $\PP(M)$ is bounded below by the translated Hodge polygon $T_1(\HP(M))$.
\end{prop}

\begin{proof}
The coefficients of $\charpoly(M)$ can be expressed as traces of exterior powers: the coefficient of $X^i$ is $\tr(\Lambda^i(M)) = \tr(\Lambda^i(U_M) \Lambda^i(\Delta_M) \Lambda^i(V_M))$.  The entries of $\Lambda^i(\Delta_M)$ can be computed easily, and now the result follows from Proposition \ref{prop:precision2}. % Clarify and expand.
\end{proof}

For many matrices, $\PP(M) = T_1(\HP(M))$.  For integer matrices with entries uniformly chosen between $0$ and $1024$, the $2$-adic precision polygon is equal to the Hodge polygon in $99.5\%$ of cases in dimension $4$, down to $99.1\%$ in dimension $8$.  For $p=3$, the fraction rises to $99.98\%$, with no clear dependence on dimension.

We may also bound $\PP(M)$ above:

\begin{prop} \label{prop:newton_prec}
If $M \in M_{n,n}(\O_K)$, the precision polygon $\PP(M)$ is bounded above by the Newton polygon $\NP(M)$.
\end{prop}
\begin{proof}
Set $f = \charpoly(M)$, set $m_{i,j}$ as the $(i,j)$th entry of $M$ and $f_{i,j}$ as the $(i,j)$th entry of $\com(X - M)$. We write $f[k]$ for valuation of the coefficient of $X^k$ in $f$, and set $f[-1] = +\infty$. The equation $(X - M) \cdot \com(X - M) = f \cdot I$  yields, for all $i$ and $k$,
\begin{align*}
f[k] &= \inf(f_{i,i}[k-1], \val(m_{i,0}) + f_{0,i}[k], \ldots, \val(m_{i,n}) + f_{n,i}[k]) \\
&\ge \inf(f_{j,i}[k]),
\end{align*}
with the infimum taken over $j$.  Taking lower convex hulls yields the result.
\end{proof}

Moreover, experiments support the following stronger results.

\begin{conj}
\begin{enumerate}
\item $\PP(M) = T_1(\HP(M))$ if $\val(\det(M))$ is $0$ or $1$, or if $M \in M_{3,3}(\O_K)$.
\item The precision polygon $\PP(M)$ is bounded above by $T_1(\NP(M))$.
\end{enumerate}
\end{conj}
Empirically, $\PP(M)$ seems most likely to differ from $T_1(\HP(M))$ at $1$, corresponding to the precision of the linear term of the characteristic polynomial.

Of course, $\PP(M)$ does necessarily agree with $\mathcal{E}$.  In $11\%$ of cases in dimension $3$, up to $15\%$ of cases in dimension $8$, the index of $\mathcal{E}$ in $\mathcal{L}_{\PP(M)}$ is greater than $1$.  This percentage increases as $\val(\det(M))$ increases, reaching $34\%$ in dimension $9$ for matrices constrained to have determinant with $2$-adic valuation $12$.

Moreover, one can construct examples with $[\mathcal{L}_{\PP(M)} : \mathcal{E}]$ arbitrarily large.  Suppose the $\sigma_i(M)$ are large.  Proposition \ref{prop:prec_hodge} implies that $\mathcal{E}$ is contained within $\O_{K, <n}[X]$ with index at least $\sum_{i=1}^{n-1} \sigma_i(M)$.  The precision lattice of $1 + M$ is obtained from $\mathcal{E}$ via the transformation $X \mapsto 1 + X$, but $\PP(1 + M)$ is now flat with height $0$.

In conclusion, for randomly chosen matrices, approximating $\mathcal{E}$ using the Hodge polygon rarely loses much precision.  However, if the $\sigma_i(M)$ are known to be large or if $M$ is a translate of such a matrix, using lattice precision can be very useful.

@@

\noindent
{\bf Numerical experiments.}
Let $B_n=(E_{i,j})_{1 \leq i,j\leq n}$ be the canonical basis of 
$M_n(K)$. It can be naturally seen as a basis of $L_n \times U_n$ as
well. For a given $M \in O_n$, let us abuse notation and write $D'(M)$
for the matrix of this linear mapping in the above basis. We remark that
$D'(M)$ is lower-triangular in this basis. Projecting $D'(M)$ onto each
coordinate, we find the best coordinate-wise loss of precision we can 
hope for the computation of $D$ is given by $\sum_u \left( \max_v \left( 
\val(D'(M)_{u,v}) \right) \right)$. This number should be compared to
$\val(\det(D'(M)))$, which is the lattice-method loss of precision.
Figure \ref{fig:LU} summarizes the mean and standard deviation of those 
losses for a sample of 2000 random matrices in $M_{d,d}(\mathbb{Z}_2)$.
%
\begin{figure}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|c|c|c|c|}
\hline 
& \multicolumn{4}{|c|}{Loss of precision in LU decomposition} \\
\cline{2-5}
\raisebox{0.2em}{matrix}
& \multicolumn{2}{|c|}{coord-wise method} & \multicolumn{2}{|c|}{lattice method}  \\  \cline{2-5}
\smash{\raisebox{0.6em}{size}} 
& \hspace{0.5em}mean\hspace{0.5em} & deviation 
& \hspace{0.5em}mean\hspace{0.5em} & deviation \\ \hline
$2$ & $\phantom{2}{-}3.0$& $5$ & $-1.5$ & $1.4$ \\% \hline
$3$& $\phantom{2}{-}9.4$& $11$ & $-2.3$ & $2.3$\\% \hline
$4$ & $-20\phantom{.4}$& $20$ & $-3.8$ & $3.1$\\ \hline
\end{tabular} 
\smallskip

@@

When the pair $(V, V^\c)$ varies, the open subsets $\mathcal 
U_{V,V^\c}$ cover the whole Grassmaniann and then define an atlas 
of it. When implementing vector spaces on a computer, we usually restrict 
ourselves to the subatlas consisting of all charts of the form $(V_I, 
V_{I^\c})$ where $I$ runs over the family of subsets of $\{1, 
\ldots, n\}$ of cardinality $d$ and $V_I$ is the subspace spanned by 
the $e_i$'s with $i \in I$. A subspace $W \in E$ then belongs to at 
least one $\mathcal U_{V_I, V_{I^{\text{c}}}}$ and, given a family of
generators of $W$, we can determine such an $I$ together with the
corresponding embedding $f : V_I \hookrightarrow E$ by echelonizing 
the matrix of generators of $W$.

@@

\noindent
{\bf A variant.}
Alternatively, one can describe $\Grass(E,d)$ as the set of linear 
surjective morphisms $f : E \to E/V$ modulo the action (by 
postcomposition) of $\GL(E/V)$ as well. This identification presents the 
tangent space at a given point $V$ as the quotient $\Hom(V, E/V) / 
\End(E/V)$, which turns out to be again isomorphic to $\Hom(V, E/V)$. 
Given a decomposition $E = V \oplus V^\c$ as above, we let $\mathcal 
U^\star_{V, V^\c}$ denote the set of surjective linear maps $f : E 
\to V^\c$ whose restriction to $V^\c$ is the identity. It is an 
affine space over $\Hom(V, V^\c)$ which can be identified with an open
subset of $\Grass(E,d)$ \emph{via} the map $f \mapsto \ker f$.

It is easily seen that $\mathcal U_{V, V^\c}$ and $\mathcal 
U^\star_{V, V^\c}$ define the same open subset in $\Grass(E,d)$. 
Indeed, given $f \in \mathcal U_{V, V^\c}$, one can write $f = \id_V 
+ h$ with $h \in \Hom(V, V^\c)$ and define the morphism $g = \id_E -
h \circ \pi \in \mathcal U^\star_{V, V^\c}$. The association $f \mapsto
g$ then defines a bijection $\mathcal U_{V, V^\c} \to \mathcal
U^\star_{V, V^\c}$ which commutes with the embeddings into the
Grassmannian.
